---
title: "Poisson EDA example (W271 Unit 5)"
author: "Michael Winton"
output: pdf_document
date: \today
header-includes:
  - \usepackage{amsmath}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)      # Ornstein dataset, Anova, residualPlots
library(effects)  # diagnostic plots for Poisson model
library(Hmisc)    # describe function
```

## Introduction

- Introduce use of a Poisson regression to model a count response variable
- Consider situations in which the conditional distribution of the response variable follow a Poisson distribution
- Using a dataset on executives at large Canadian companies, study interocking directorates (board members serving on each other's boards)

## Dataset

This example works with the `Ornstein` dataset from the `car` library

```{r}
# list the top 10 observations in the dataset
head(Ornstein, 10)

# summary statistics about the dataset
summary(Ornstein)
describe(Ornstein)  #more verbose summary stats

# show mean value of interlocks per nation|sector
round(with(Ornstein, tapply(interlocks, nation, mean)), 1)
round(with(Ornstein, tapply(interlocks, sector, mean)), 1)
```

In the original study, the author performed an OLS regression with the response variable being `interlocks`, the number of interlocks maintained by each firm, on firm's assets (in millions of dollars), sector of operation, and nation of control.

However, as the variable `interlocks` is a count, a *Poisson* regression model may be more appropriate.

Let's take a look at the distribution of the `interlocks` variable.  We will have to construct this graph in multiple steps.

## Analyis of the Response Variable
```{r}
# Frequency distribution of the interlocks
tab <- xtabs(~interlocks, data=Ornstein)
str(tab)      # this view is less useful for a table
class(tab)    # check class (so we know what methods and attributes are available)
names(tab)    # variable names
nrow(tab)
tab

# Record the distinct values of the interlocks
x <- as.numeric(names(tab))  # xtabs stored these as char variables

# Scatter plot of data
plot(x, tab, xlab='Number of Interlocks', ylab='Frequency')

# Bar plot of frequencies
plot(x, tab, type='h',  xlab='Number of Interlocks', ylab='Frequency')
points(x, tab, pch=16)  # decorate top of bars with a point
```

## Quick Analysis of Exploratory Variables

```{r}
# Histogram (untransformed)
with(Ornstein, hist(assets, breaks='FD', col='navy', main='Distribution of Assets, millions'))
```

Shape suggests a log transform may help the distribution be more approximately normal.

```{r}
# Histogram (log2 transformed)
with(Ornstein, hist(log2(assets), breaks='FD', col='navy', main='Distribution of Log_2(Assets), millions'))

```

```{r}
# scatterplot matrices to look at variable correlations
scatterplotMatrix(~assets + nation + sector, data=Ornstein)
scatterplotMatrix(~log2(assets) + nation + sector, data=Ornstein)
```

*Note: this is NOT a good approach.*  Scatterplot matrices don't make a lot of sense in this case since explanatory variables are categorical.  SCM doesn't display the features very well.  Instead, use boxplots:

```{r}
Boxplot(assets ~ nation, data=Ornstein, main='Assets by Nation')
Boxplot(assets ~ sector, data=Ornstein, main='Assets by Sector')
```

*Observations:* Because `assets` is extremely right-skewed, we see the extended tail of points for `CAN`.   Banking sector has the largest assets, and also extremely wide distribution. Since data is extremely right-skewed, we may want to look at modified plots: 95th percentile assets; excluding banking.

```{r}
summary(Ornstein$assets)
(pctile95 <- quantile(Ornstein$assets, 0.95))

Boxplot(assets[assets<pctile95] ~ nation[assets<pctile95], data=Ornstein, main='Assets by Nation (Assets < ~$20B)')
Boxplot(assets[sector!='BNK'] ~ sector[sector!='BNK'], data=Ornstein, main='Assets by Sector (Excluding Banking)')

```

## Poisson Regression

Now, build the regression model.  Note that we're using the log-transformation of `assets`.

```{r}
poisson_fit <- glm(interlocks ~ log2(assets) + nation + sector, family=poisson(link=log), data=Ornstein)
summary(poisson_fit)
```

*Observations:* We can look at the summary to see which are the (omitted) base cases.  In this case, it's `nation=CAN` and `sector=AGR`.   Also, remember that the coefficients are for the *linear predictor*, ie. for the *log*(counts).  We have to exponentiate to get effect on the counts.

We can also pull out residual device from the model and calculate a "goodness of fit" p-value.  The deviance and degrees of freedome come from the fit:

```{r}
resid_dev <- with(poisson_fit, cbind(resid_deviance = deviance, df = df.residual,
                                     p = pchisq(deviance, df.residual, lower.tail = FALSE)))
round(resid_dev, 4)
```
Because the goodness of fit $\chi^2$ test is statistically significant, we can conclude the model does not fit the data particularly well.

Let's look at the frequency table.

```{r}
# Create frequency table
with(Ornstein, table(nation, sector))
```

*Observations:* Our reference category (CAN + AGR) has a substantial number of observations.  This is good; if our base case had a small number, we might be better off changing the base case.

** Analysis of Deviance

```{r}
Anova(poisson_fit)
```

*Observations:* Capital `Anova` does a Type II analysis.  All predictors have high significance.

## Interpretation of Coefficients

Coefficients of the model are interpreted as effects on the log-count scale (ie. the scale of the linear predictor). Exponentiating the coefficients produces the multiplicative effects on the count scale.

```{r}
exp(coef(poisson_fit))
```

*Observations:* a corporation that is twice the size (based on assets) of another one has an estimated 36.7% higher number of interlocks, holding all other factors constant.

A US firm on average maintains only 46.2% as many interlocks as a Canadian firm, holding all other factors constant.

## Visualize the Effects of Changes in Explanatory Variables

Effects plots show how estimated responses change with predictor values.  (These are not diagnostic plots.)  

One variable at a time is changed, while others are held constant at "typical" values.  (For continuous variables, at their mean; for factor variables, sets proportional distribution to match that observed in data.)  

```{r}
plot(allEffects(poisson_fit, default.levels=50))
```

The y-axis is on the scale of the linear predictor (in this case, `log`).  The bands around the means are 95% confidence intervals, calculated using standard errors from the model.   From the y-axis range of data, we can get a sense of strength of each predictor's effect; we can also get a sense of level of uncertainty in the estimates.

## Model Diagnostics

The diagnostic plots for a Poisson model are different than for linear regression.  These plots use Pearson residuals.  There is one plot per explanatory variable, as well as one for the linear predictor.  We want to see that there are _no systematic patterns_ between residuals and the explanatory variables.  We should look for nonlinear trends (esp. curved splines), trends in variation across the graph, and outlier points.  Box plots should show roughly similar centers and spreads.

```{r}
residualPlots(poisson_fit, layout=c(2,2))
```

*Observations:* not much pattern observed here.  The *lack of fit test* results are also displayed (only meaniningful for continuous variables, otherwise `NA`).  According to the R help text, this is a curvature test.  For plots against a term in the model formula, say $X$, the test displayed is the t-test for for $I(X^2)$ in the fit of an updated model: ~ . + $I(X^2)$.

The null hypothesis of this test is $H_0:$ the model fits the data well (ie. the coefficient for a curvature term $\beta_{I(X^2)}= 0$), vs. $H_a:$ the coefficient for a curvature term is non-zero.  _Here, the p-value of ~0 indicates a lack of fit, even thought it's not as evident from visual inspection._

We can also look for influential points:

```{r}
influenceIndexPlot(poisson_fit, vars = "Cook")
influenceIndexPlot(poisson_fit, vars = "hat")
```

