---
title: "Intro to Time Series"
author: "Michael Winton"
date: "6/16/2018"
output:
  html_document:
    df_print: paged
header-includes: \usepackage{amsmath}
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=4)
library(car)
library(dplyr)
```

## Introduction

Observations that have been collected over fixed sampling intervals are _historical time series_.  They are treated as realizations of sequences of random variables.  

```{r}
# load sample dataset
data(AirPassengers)
(ap <- AirPassengers)
class(ap)  # Note there's a specific class called ts

start(ap)
end(ap)
frequency(ap)

# rollups by year
aggregate(ap)

```

Several common plots are built in:
```{r}
# raw time series
plot(ap, ylab='Passengers (1000s)')

# annual total counts
plot(aggregate(ap), lty=2, ylab='Annual Total Passengers (1000s)')

# box plots by month
boxplot(ap ~ cycle(ap), ylab='Passengers (1000s)')
```

The simplest model for a (long-term) _trend_ is often a linear approximation.

Converting dataframe to time series:

```{r}
# load data from internet and transform to a ts object
url <- 'https://raw.githubusercontent.com/mwinton/Introductory_Time_Series_with_R_datasets/master/Maine.dat'
ME_month_df <- read.table(url, header=TRUE)
head(ME_month_df)
(ME_month_ts <- ts(ME_month_df$unemploy, start=c(1996, 1), freq=12))
(ME_annual_ts <- aggregate(ME_month_ts, FUN=mean))  # without FUN=mean, we get a sum by default

plot(ME_month_ts, ylab='Unemployed (%)')
plot(ME_annual_ts, ylab='Unemployed (%)')
```

We can extract just a portion of the ts:
```{r}
# Extract all February and August numbers
(ME_feb <- window(ME_month_ts, start=c(1996,2), freq=TRUE))
(ME_aug <- window(ME_month_ts, start=c(1996,8), freq=TRUE))
# Compare Feb, Aug to average monthly unemployment (from entire ts)
(feb_ratio <- mean(ME_feb) / mean(ME_month_ts))
(aug_ratio <- mean(ME_aug) / mean(ME_month_ts))
```

## Multiple time series:
```{r}
url <- 'https://raw.githubusercontent.com/mwinton/Introductory_Time_Series_with_R_datasets/master/cbe.dat'
cbe_df <- read.table(url, header=TRUE)
head(cbe_df)

choc_ts <- ts(cbe_df$choc, start=1958, freq=12)
beer_ts <- ts(cbe_df$beer, start=1958, freq=12)
elect_ts <- ts(cbe_df$elec, start=1958, freq=12)

# cool trick to use cbind to join plots together on shared t axis
plot(cbind(choc_ts, beer_ts, elect_ts))

# intersection of multiple time series
isect_ap_elec <- ts.intersect(ap, elect_ts)
start(isect_ap_elec)
end(isect_ap_elec)
head(isect_ap_elec)

# plot both
layout(1:2)
plot(isect_ap_elec[,2], ylab="Airline Passengers")
plot(isect_ap_elec[,1], ylab="AUS Electricity Production")

# need to convert to vectors to do scatterplot
layout(1:1)
plot(as.vector(isect_ap_elec[,1]), as.vector(isect_ap_elec[,2]),
     xlab="AUS Electricity Production", ylab="Airline Passengers")
abline(lm(as.vector(isect_ap_elec[,2]) ~ isect_ap_elec[,1]), lty=2)
```

From this example, we see how easy it is to find correlation in two completely unrelated things.  Because of this, its common to remove _trends_ and _seasonal variation_ before doing multiple time series analysis.  This could mean working with residuals from a regression model that has terms to represent them.

## Stochastic Trends in Financial Data

Day-to-day changes can be hard to explain, and it may be unrealistic to assume any _deterministic_ component in a time series model.  _Stochastic trends_ can sometimes be fit with _random walk_ models.  They're common in financial data.  There are also statistical tests to test for stochastic trends.

```{r}
url <- 'https://raw.githubusercontent.com/mwinton/Introductory_Time_Series_with_R_datasets/master/pounds_nz.dat'
exchange_df <- read.table(url, header=TRUE)
head(exchange_df)
exchange_ts <- ts(exchange_df$xrate, start=1991, freq=4)
plot(exchange_ts, xlab='Quarter', ylab='Exchange Rate (NZ$ to GBP)')
```

```{r}
url <- 'https://raw.githubusercontent.com/mwinton/Introductory_Time_Series_with_R_datasets/master/global.dat'
global_df <- scan(url)
head(global_df)
global_ts <- ts(global_df, start=c(1856, 1), end=c(2005, 12), freq=12)
# create ts for average annual value
global_annual <- aggregate(global_ts, FUN=mean)

# plot time series
layout(1:2)
plot(global_ts)
plot(global_annual)

# plot subset of our ts
global_recent_ts <- window(global_ts, start=c(1970, 1), end=c(2005, 12))
recent_time_axis <- time(global_recent_ts)  # for regression
layout(1:1)
plot(global_recent_ts, ylab='Temp Increase (C)')
abline(reg=lm(global_recent_ts ~ recent_time_axis), lty=2)
```